Comparison of Terrain Following and Cut Cell Grids with a Non-Hydrostatic Model

### 2015-04-10 msc-publication.tex
more discussion of Lorenz computational mode, add introduction to sections, introduce grids section

This study uses the nonhydrostatic model from \citet{weller-shahrokhi2014} to enable a like-for-like comparison between BTF, SLEVE and cut cell grids for idealised, two-dimensional test cases from the literature.  \TODO{final bit of intro: in section 2, section 3 etc etc}

@@

\TODO{introduce this section}
\citet{galchen-somerville1975} proposed a basic terrain following (BTF) coordinate defined as 
\begin{equation}
	z = \left( H - h \right) \left( z^\star / H \right) + h \label{eqn:btf}
\end{equation}
where, in two dimensions, \(z(x, z^\star)\) is the height of the coordinate surface at level \(z^\star\), \(H\) is the height of the domain, and \(h(x)\) is the height of the terrain surface.  Using this coordinate, the terrain's influence decays linearly with height but disappears only at the top of the domain.

@@

In a further test, the mountain height is doubled from \SI{250}{\meter} to \SI{500}{\meter} with all other parameter values unchanged.  The same spurious anomaly in potential temperature is again present on the cut cell grid but its amplitude increases.  Figure~\ref{fig:gw-exner-theta} shows vertical profiles of the Exner function of pressure and potential temperature in the lowest \SI{1}{\kilo\meter} in the lee of the mountain at \(x = \SI{50}{\kilo\meter}\).

Analysis:
\begin{itemize}
	\item \TODO{divergence might help explain computational mode -- it's greater on the SnapCol grid (MSc dissertation Figure 4.15e)}
	\item \TODO{worth mentioning implications of this?  1. reduced stability, although not enough to create vertical motion in this instance 2. would disrupt clouds 2. model has no viscosity so thermal mixing should not occur in theory}
	\item Lorenz computation mode has been excited because Exner sample line at $x = \SI{50}{\kilo\meter}$ is in hydrostatic balance (Figure~\ref{fig:gw-exner-theta})
	\item Conclusion: results similar on all grids, agree with literature, except for Lorenz computational mode on SnapCol grid
\end{itemize}

>>>>

This study uses the nonhydrostatic model from \citet{weller-shahrokhi2014} to enable a like-for-like comparison between terrain following and cut cell grids for idealised, two-dimensional test cases from the literature.  Section 2 presents the formulation of the terrain following and cut cell grids used in the experiments that follow.  Section 3 analyses the results from two tracer advection tests, a test of a stratified atmosphere at rest, and orographically induced gravity waves.  Concluding remarks are made in section 4.

@@

In this section we describe the formulation of the terrain following grids and the method of cut cell grid construction.  The techniques presented here are used to define the grids for the experiments in the subsequent section.

\citet{galchen-somerville1975} proposed a basic terrain following (BTF) coordinate defined as 
\begin{equation}
	z = \left( H - h \right) \left( z^\star / H \right) + h \label{eqn:btf}
\end{equation}
where, in two dimensions, \(z(x, z^\star)\) is the height of the coordinate surface at level \(z^\star\), \(H\) is the height of the domain, and \(h(x)\) is the height of the terrain surface.  Using this coordinate, the terrain's influence decays linearly with height but disappears only at the top of the domain.

@@

In a further test, the mountain height is doubled from \SI{250}{\meter} to \SI{500}{\meter} with all other parameter values unchanged.  The same spurious anomaly in potential temperature is again present on the cut cell grid but its amplitude increases.  Figure~\ref{fig:gw-exner-theta} shows vertical profiles of the Exner function of pressure and potential temperature in the lowest \SI{1}{\kilo\meter} in the lee of the mountain at \(x = \SI{50}{\kilo\meter}\).  We find that the Exner function of pressure decreases linearly with height on all grids.  The potential temperature increases linearly with height on the BTF and SLEVE grids.  On the cut cell grid, the potential temperature anomalies seen in figure~\ref{fig:gw-theta}d appear as a zig--zag in figure~\ref{fig:gw-exner-theta}.

The model by \citet{weller-shahrokhi2014} has a Lorenz staggering of variables so that $\theta$ is stored at cell centres and velocity at cell faces.  The inaccuracies in potential temperature become invisible when calculating momentum because $\theta$ is interpolated onto cell faces, so discrete hydrostatic balance is maintained.  This is a manifestation of the Lorenz computational mode, which is a disadvantage of the Lorenz grid that has been widely discussed in the literature \citep{arakawa-konor1996,holdaway2013b}.

\begin{itemize}
	\item \TODO{worth mentioning implications of this?  1. reduced stability, although not enough to create vertical motion in this instance 2. would disrupt clouds 2. model has no viscosity so thermal mixing should not occur in theory}
	\item Conclusion: results similar on all grids, agree with literature, except for Lorenz computational mode on SnapCol grid
\end{itemize}
